### 2017-01-03 introduction.tex
Rewrote observations of polar structuring section

@@

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here we will focus on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  The first observations of polar patches were made in the 1960s by \citet{Hill1963}.  They can range in size from 100 km up to 1500 km in diameter and can have peak densities as much as eight times that of the background ionospheric plasma \citep{Weber1986,Hosokawa2014}.  Polar patches are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \citep{Weber1985,Weber1986}.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  The spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981,Keskinen1982,Gondarenko2001,Gondarenko2004b}.  However, very strong shears or increasing the impact of ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.

@@

%\citet{Keskinen1982} found plasma waves on the trailing edge of a large density enhancement are unstable.  \citet{Keskinen1990} expanded these results by considering coupling between the magnetosphere and ionosphere, but found that this coupling only has a substantial effect for very large-scale structuring.  In addition, this simulation predicted instability evolution that was far faster than what observations had shown.  3D simulations of GDI surrounding a plasma density enhancement were introduced in \citet{Guzdar1998}.  \citet{Gondarenko1999} improved this 3D simulation further by including plasma dynamics parallel to the magnetic field and inertial effects.  These factors are necessary for the simulation to model secondary KHI and tertiary shear-driven instability processes.  These simulations provided insight in how exactly GDI structuring develops.

@@

\subsection{Observations of Polar Cap Structuring}
\label{sec:lit_observations}
Polar patches were first observed in the 1960s \citep{Hill1963}, but much of the work of characterizing them using optical and radio techniques was not accomplished until two decades later \citep{Weber1981,Weber1984,Weber1986,Buchau1983,Buchau1985}.  \citet{Weber1981} identified sun-aligned arcs that extend  along the sun-earth line through large parts of the polar cap using all-sky imaging photometers (ASIP).  These arcs are over 1000 km long and \(\sim100\) km wide and are produced by soft electron precipitation.  \citet{Buchau1983} used a combinations of Digisonde and ASIP measurements to identify large luminous patches in the polar cap in addition to sun-aligned arcs.  These patches could have densities up to \(10^6\) cm\(^{-3}\) and tended to drift antisunward.  Additionally, \citet{Buchau1983} found that polar patches were more likely to be observed for southward IMF \(B_z\) while sun-aligned arcs were more common for northwards \(B_z\).  \citet{Weber1984} found patches of enhanced ionization between 800--1000 km during moderately disturbed geomagnetic conditions, again using a combination of ASIP images, ionosonde measurements, and data from Dynamic Explorer 2 (DE-2).  These findings confirmed that density enhancements tended to drift antisunward at speeds of 500-1000 m/s and patches were not colocated with enhanced particle precipitation within the polar cap, indicating that this is not the source of the density enhancement.  Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge.  Conversely, \citet{Buchau1985} found that plasma densities within patches were comparable with those measured at dayside sub-auroral latitudes, suggesting that dayside solar-illuminated plasma is the source of dense plasma that form polar patches.

\citet{Coley1998} examined the occurrence of polar patches in both the northern and southern hemispheres with DE-2 and found that the frequency of patch occurrence was greatest when  the cusp was slightly dayward of the terminator.  This creates the situation of an unsunlit polar cap, so background plasma density is low and blobs of enhanced density are easily visible.  In addition, the cusp then provides a gateway of highly ionized sunlit plasma into the dark polar cap.  Several studies had previously reported density enhanced patches convecting from the dayside through the cusp/throat region toward the polar cap \citep{Kelly1984,Foster1984,Foster1985,Foster1993}.  This also agrees with observations made by \citet{Sojka1982,delaBeaujardiere1985}  \citet{Rodger1996} also considered occurrence of patches but with a HF radar in Halley, Antarctica.  The diurnal variation in patch occurrence was found to peak at magnetic noon and the seasonal variation in equinox months.  The only correlation between hourly averaged solar wind parameters that was found was a bias towards patches occurring with a negative IMF \(B_z\) component.  A review of historical observations of polar patches has been provided by \citet{Crowley1996}

The theoretical mechanism for patch formation via transient magnetic reconnection has been described previously, Section \ref{sec:patch_formation}.  A variety of experimental evidence has been presented to support this mechanism \citep{Cowley1998,Carlson2002}.  Experiments run in the 1980s using the European Incoherent Scatter Scientific Association (EISCAT) Troms\o, Norway ISR showed that plasma flow occurred near noon shortly after southward IMF \(B_z\) arrives at the magnetopause \citep{Etemadi1988,Todd1988}.  Later, \citet{Lockwood1993a,Lockwood1993b} showed that these plasma flows are pulsed in the region of the polar cap boundary, as predicted by \citet{Cowley1991}.  Observations made by an ASIP in Svalbard in 1984 are fully consistent with patch formation through transient magnetopause reconnection \citep{Carlson1996,Carlson2002}.  \citet{Carlson2004} identified five signatures of transient magnetic reconnection and then examined a patch formation event for these signatures using the EISCAT Svalbard Radar.  All five signatures were observed as expected, strengthening the idea that transient magnetopause reconnection is responsible for patch formation.  \citet{Carlson2006} furthered this using a observations from the EISCAT Svalbard Radar, the EISCAT VHF radar in Troms\o an ASIP, and a meridian scanning photometer (MSP).  They managed to identify and track a series of patches directly from the subauroral plasma reservoir and show that the boundary moves equatorward before relaxing poleward, one of the main predictions of the transient magnetopause reconnection mechanism \citep{Lockwood1992b}.

@@

This asymmetry was first observed by \citet{Weber1984}.  Backscatter power from HF radars tends to be greater on the trailing edges of both moving high-density polar structures \citep{Milan2002b} and sun-aligned arcs identified from ASIP data \citep{Koustov2012}, which is generally attributed to GDI being unstable on the trailing edge but stable on the leading edge of such large, moving density structures.  Studies of patches using all-sky airglow imagers (ASI) found that the density gradient on the leading edge of polar patches tends to be 2--3 times steeper than that on the trailing edge \citep{Hosokawa2016}.  Additionally, large finger-like structures on the trailing edge of patches have been identified that were 10s--100s of kilometers in size and agree with predictions made by GDI simulations \citep{Gondarenko2004b}.  \citet{Moen2012} made direct measurements of plasma density structuring using the ICI-2 sounding rocket.  This study provided evidence that decameter scale plasma structuring (the same as observed by HF radars) had spawned from kilometer scale density gradients (that observed by ASIP and ASI methods) and structuring was greatest where GDI was operational.

>>>>

The majority of mass in the universe exists as plasmas.  Plasmas are fluid masses of charged particles, often formed when neutral atoms and molecules in a gas are ionized by external energy.  Everyday examples of this is florecent lightbulbs, as well as flames.  More exotic instances of plasma include nuclear fusion reactor experiments, most stars, nebula, and the interstellar medium.  In fact, about 99\% of matter in the known universe exists in the plasma state.  Of particular interest to this thesis is the plasma in Earth's ionosphere.

Plasmas can be either fully ionized, where most particles in the plasma are ionized, or partially ionized, where only some fraction are ionized and the rest are neutral.  Considering the dynamics of plasmas therefore requires not only collisions between particles to be considered but also charge-charge interactions and external electric and magnetic fields.  These dynamics can generate waves in the plasma, which cause density perturbations.  Additionally, variations in temperature and electric and magnetic field strength can occur on a variety of scales.  This thesis focuses on mechanisms by which density variations in ionospheric plasma occur, often referred to as plasma structuring.

@@

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here we will focus on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  One of the primary interests of this thesis is the relationships between irregularities of different scales and small-scale structuring has long been observed around polar patches \citep{Weber1984,Milan2002,Moen2012}.  Although small-scale structures have also been observed around sun-aligned arcs \citep[e.g.][]{Koustov2012}, these structures also tend to be more complicated and often are surrounded by intense shears and electric fields \citep{Safargaleev2000,Aikio2002,Kozlovsky2007}.  Additionally, polar patches occur relatively frequently, providing a good opportunity to study irregularity coupling at different scales \citep{Rodger1996}.  In theory, many of the results presented in this work pertaining to the strong gradients on the edges of patches could also easily be applied to the edges of polar holes, but to date, far less attention has been paid to polar holes in the literature, making it challenging to accurately put observations in context \citep{Makarevich2015}.

The first observations of polar patches were made in the 1960s by \citet{Hill1963}.  They can range in size from 100 km up to 1500 km in diameter and can have peak densities as much as eight times that of the background ionospheric plasma \citep{Weber1986,Hosokawa2014}.  Polar patches are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \citep{Weber1985,Weber1986}.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  The spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981,Keskinen1982,Gondarenko2001,Gondarenko2004b}.  However, very strong shears or increasing the impact of ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.  The power spectra derived from these simulations are similar to those from experimental observations \citep{Baker1978,Kelley1979}, however ... \textbf{WHAT'S DIFFERENT AND IMPORTANT ABOUT PROJECT 3 HERE?}

@@

%\citet{Keskinen1982} found plasma waves on the trailing edge of at large density enhancement are unstable.  \citet{Keskinen1990} expanded these results by considering coupling between the magnetosphere and ionosphere, but found that this coupling only has a substantial effect for very large-scale structuring.  In addition, this simulation predicted instability evolution that was far faster than what observations had shown.  3D simulations of GDI surrounding a plasma density enhancement were introduced in \citet{Guzdar1998}.  \citet{Gondarenko1999} improved this 3D simulation further by including plasma dynamics parallel to the magnetic field and inertial effects.  These factors are necessary for the simulation to model secondary KHI and tertiary shear-driven instability processes.  These simulations provided insight in how exactly GDI structuring develops.

@@

\subsection{Observations of Polar Cap Structuring}
\label{sec:lit_observations}
Polar patches were first observed in the 1960s \citep{Hill1963}, but much of the work of characterizing them using optical and radio techniques was not accomplished until two decades later \citep{Weber1981,Weber1984,Weber1986,Buchau1983,Buchau1985}.  Polar patches are density enhancements of as much as \(10^6\) cm\(^{-3}\) that occur primarially in the polar F region \citep{Buchau1983}.  There were initially catagorized as extending 800--1000 km horizontally \citep{Weber1984}, but more recent studies have identified patches as large as 1500 km \citep{Hosokawa2014}.  Patches tend to drift antisunwards at speeds of 500--1000 m/s, following background convection \citep{Buchau1983,Weber1984}.

%\citet{Coley1998} examined the occurrence of polar patches in both the northern and southern hemispheres with DE-2 and found that

Polar patch occurrence varies both diurnally and seasonally, with a diurnal peak at magnetic noon and a seasonal peak during equinox months \citep{Rodger1996}.  Additionally, there is a bias towards patches occurring when there is a negative IMF \(B_z\) component \citep{Buchau1983,Rodger1996}.  The frequency of patch occurrence is greatest when the cusp is slightly dayward of the terminator.  This creates the situation of an unsunlit polar cap, so background plasma density is low and blobs of enhanced density are easily visible \citep{Coley1998}.  In addition, the cusp then provides a gateway of highly ionized sunlit plasma into the dark polar cap.  Patches convect from the dayside through the cusp/throat region towards the polar cap \citep{Kelly1984,Foster1984,Foster1985,Foster1993,Sojka1982,delaBeaujardiere1985}.

%Several studies had previously reported density enhanced patches convecting from the dayside through the cusp/throat region toward the polar cap \citep{Kelly1984,Foster1984,Foster1985,Foster1993}.  This also agrees with observations made by \citet{Sojka1982,delaBeaujardiere1985}

%\citet{Rodger1996} also considered occurrence of patches but with a HF radar in Halley, Antarctica.  The diurnal variation in patch occurrence was found to peak at magnetic noon and the seasonal variation in equinox months.  The only correlation between hourly averaged solar wind parameters that was found was a bias towards patches occurring with a negative IMF \(B_z\) component.

Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge.

\citet{Weber1981} identified sun-aligned arcs that extend  along the sun-earth line through large parts of the polar cap using all-sky imaging photometers (ASIP).  These arcs are over 1000 km long and \(\sim100\) km wide and are produced by soft electron precipitation.

Additionally, \citet{Buchau1983} found that polar patches were more likely to be observed for southward IMF \(B_z\) while sun-aligned arcs were more common for northwards \(B_z\).

%\citet{Buchau1983} used a combinations of Digisonde and ASIP measurements to identify large luminous patches in the polar cap in addition to sun-aligned arcs.  These patches could have densities up to \(10^6\) cm\(^{-3}\) and tended to drift antisunward.

%\citet{Weber1984} found patches of enhanced ionization between 800--1000 km during moderately disturbed geomagnetic conditions, again using a combination of ASIP images, ionosonde measurements, and data from Dynamic Explorer 2 (DE-2).  These findings confirmed that density enhancements tended to drift antisunward at speeds of 500-1000 m/s and patches were not colocated with enhanced particle within the polar cap, indicating that this is not the source of the density enhancement.  Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge.

Patches are not colocated with enhanced particle precipitation within the polar cap, indicating that this is not the source of the density enhancement \citep{Weber1984}.  Conversely, \citet{Buchau1985} found that plasma densities within patches were comparable with those measured at dayside sub-auroral latitudes, suggesting that dayside solar-illuminated plasma is the source of dense plasma that form polar patches.

The theoretical mechanism for patch formation via transient magnetic reconnection has been described previously, Section \ref{sec:patch_formation}.  A variety of experimental evidence has been presented to support this mechanism \citep{Cowley1998,Carlson2002}.  Experiments run in the 1980s using the European Incoherent Scatter Scientific Association (EISCAT) Troms\o, Norway ISR showed that plasma flow occurred near noon shortly after southward IMF \(B_z\) arrives at the magnetopause \citep{Etemadi1988,Todd1988}.  These plasma flows are pulsed in the region of the polar cap boundary \citep{Lockwood1993a,Lockwood1993b}, as predicted by \citet{Cowley1991}.  Observations made by an ASIP in Svalbard in 1984 are fully consistent with patch formation through transient magnetopause reconnection \citep{Carlson1996,Carlson2002}.  \citet{Carlson2004} identified five signatures of transient magnetic reconnection and then examined a patch formation event for these signatures using the EISCAT Svalbard Radar.  All five signatures were observed as expected, strengthening the idea that transient magnetopause reconnection is responsible for patch formation.  Furthermore, \citet{Carlson2006} identified and tracked a series of patches directly from the subauroral plasma reservoir and show that the boundary moves equatorward before relaxing poleward, one of the main predictions of the transient magnetopause reconnection mechanism \citep{Lockwood1992b}.  A review of historical observations of polar patches has been provided by \citet{Crowley1996}.

@@

Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge \citep{Weber1984}.  Backscatter power from HF radars tends to be greater on the trailing edges of both moving high-density polar structures \citep{Milan2002b} and sun-aligned arcs identified from ASIP data \citep{Koustov2012}, which is generally attributed to GDI being unstable on the trailing edge but stable on the leading edge of such large, moving density structures.  Studies of patches using all-sky airglow imagers (ASI) found that the density gradient on the leading edge of polar patches tends to be 2--3 times steeper than that on the trailing edge \citep{Hosokawa2016}.  Additionally, large finger-like structures on the trailing edge of patches have been identified that were 10s--100s of kilometers in size and agree with predictions made by GDI simulations \citep{Gondarenko2004b}.  \citet{Moen2012} made direct measurements of plasma density structuring using the ICI-2 sounding rocket.  This study provided evidence that decameter scale plasma structuring (the same as observed by HF radars) had spawned from kilometer scale density gradients (that observed by ASIP and ASI methods) and structuring was greatest where GDI was operational.

Small-scale plasma irregularities are typically studied with HF coherent scatter radars in the polar cap, Section \ref{sec:csr}.  The factors that affect irregularity observation can roughly be divided into two groups, irregularity production and radar propagation.  Irregularity production factors are related to the existance of plasma irregularities in the ionosphere while radar propagation factors are related to the abililty to observe irregularities with a ground based radar, Section \ref{sec:superdarn}.  Within the polar caps, plasma irregularity production is much higher than at lower altitudes such that when background plasma conditions are favorable for radar propagation, FAIs are observed nearly continuously \citep{Bristow2011}.  However, there are still definite diurnal, seasonal, and solar cycle trends \citep{Kane2012}.  There is an increase in echo occurrence at solar maximum, particularly in the midnight sector \citep{Milan1997,Koustov2004}.  Seasonally, high latitude radars observe a peak in echo occurrence durring equinox months \citep{Koustov2004}.  Additionally, backscatter tends to occure at closer ranges in summer at midnight, but moves further away from the radar in other seasons at midday \citep{Milan1997}.  In general, echo occurrence in high durring the day and low at night, but nighttime occurrence is directly proportional to the F peak density and daytime occurrence can be significantly reduced by a dense and conductive E layer. \citep{Koustov2004,Kane2012,Vickrey1982}.  In the daytime, solar illumination causes the electron density of the E region to increase dramatically, making it highly conductive.  This conductive layer can provide a conduit by which the potential differences associated with FAIs in the F region can be shorted out, effectively reducing F-region plasma irregularities \citep{Vickrey1982}.  Both irreguarity production and radar propagation factors probably contribute to these regular variations of echo occurrence, because while a dense, daytime ionosphere creates favorable radar propigation conditions, it can also contribute to E-region shorting or smoothing of gradients which surpress irregularity production \citep{Koustov2004}.

Irregularity production is impacted primarially by the presence of strong density gradients and electric fields in the ionosphere \citep{Koustov2004}.  Properly oriented density gradients and electric fields are a nessesary condition for GDI to be operational, Section \ref{sec:lit_instbilities}, so the stronger these are, the more likely the irregularity production rate is high.  Both backscatter power and occurrence increase with increasing electric field strength \citep{Ballatore2001,Danskin2002,Makarevich2014}.  Similarly, backscatter occurrence increases when the IMF \(B_z\) component is strongly negative, implying a strong merging electric field.

The primary radar propogation factor to consider is the background plasma density and how much the radar beam is refracted.  If the background density is too low, the beam is under refracted and is never perpendicular to the magnetic field, however if the density is too high, the beam is over refracted, such that observation of small-scale FAIs is most favorable if the background density is between \(1\times 10^{11}\) m\(^{-3}\) and \(4\times 10^{11}\) m\(^{-3}\) \citep{Danskin2002,Makarevich2014}.  If the D region is too dense, this can also in theory attenuate the signal and reduce the rate of returned backscatter, although observations have shown this is not a major factor in echo occurrence \citep{Danskin2002}.  Finally, the radar equation predicts that the returned backscatter should be proportional to the perturbation density, \(\delta n\), squared, or, equivilently, the product of the background density and the fractional density, \(\delta n/n\), squared, Equation \ref{eqn:fractional_density}
\begin{equation}
	\label{eqn:fractional_density}
	P \propto \delta n^2 \propto \left(\frac{\delta n}{n}\right)^2 n^2
\end{equation}.
There is no general agreement of what controls the fractional density, however, if it is approximately constant, then radar backscatter power could be directly proportional to \(n^2\).

%Kane2012:
%- diurnal, seasonal, solar cycle variation
%- nighttime occurrence dirrectly proportional to peak electron density
%- daytime occurrence significantly reduced by conductive E layer
%Danskin2002:
%- D region absorption not a major factor in echo occurrence
%- under refraction occurs if n<1x10^11 m^-3
%- over refraction occurs if n>4x10^11 m^-3
%- echo power increases for stronger electric fields
%Bristow2011:
%- polar cap has nearly continuous production of decameter scale irregularities
%- "near constant presnece of field-aligned density irregularities in the polar cap"
%Milan1997:
%- summer - backscatter observed at near ranges in midnight sector
%- winter/equinox - backscatter observed at far ranges in midday sector
%- frequency/boresite azimuth have a large impact on the ability to detect close backscatter
%- increased backscatter at solar max
%Makarevich2014:
%- power increases with increasing density and electric field
%- power drops significantly at large density values
%- backscatter power proportional to n^2
%Koustov2013:
%- antisunward flows occur for negative Bz, sunward for positive
%Koustov2004:
%- increasced echo occurrence for solar max in midnight sector due to strong density gradients and electric fields
%- noon sector does not see occurance increase with solar max
%- high latitude radars have occurrence max in equinox months
%- combination of propagation and production factors
%Ballatore2001:
%- backscatter occurrence increases for negative Bz
%- backscatter occurrence increases with merging electric field strength
